Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-08-05 paper/manuscript.tex
Includes changes from @hannahbruce discussed in https://github.com/alchemistry/alchemical-best-practices/pull/162 which partially address https://github.com/alchemistry/alchemical-best-practices/issues/141.
Other tweaks also included.

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
\end{itemize}

@@

\paragraph{Have you considered stereoisomers and enantiomers?}
Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known. Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. However, the modeling typically uses just the bioactive conformation that best fits the active site. Clearly this introduces potential for larger errors compared to experiment. Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.

@@

\subsection{Free energies of hydration or partition coefficients}
\tocsubsectioncomment{Steering Meeting: @Volunteer to expand coverage of free energies of hydration/ partition coefficients to make coverage of applications more equal.}
\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. However, in these cases a careful choice of forcefield for the organic solvent model, as well as water model is essential. See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices. And, while sampling problems might seem to be a non-issue for small molecules, this is not always the case; e.g. even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{klimovich2010predicting, lim2019assessing}.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details. In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond. The other energy terms are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.

The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$. Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems. Giese et al. \cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors. \cite{ries2024Kartograf}

@@

In general there are mutliple ways on how a MCSS can be calculated. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry. Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}. The SIS MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.

Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules. These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g. stereochemistry or binding mode differing).
The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned. Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.

In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

@@

While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}. Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule. However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom. Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al. referred to this as “flapping”. Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed. In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small. For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017, bieniek2021}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task. In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}. In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them. Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses. However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues. Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear. Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.

@@

\subsection{Uncertainty estimation}
\tocsubsectioncomment{To be split between single/ multiple papers. @Agastya P Bhati/ Peter Coveney to expand. Cycle closure etc. to be moved to second paper.}
\tocsubsectioncomment{@Volunteer: Add discussion of bootstrapping vs t-based confidence intervals for repeats of single calculations (both bad options)?}
\label{subsec:uncertainty}
It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to obtain an estimate of uncertainty of the obtained value for the free energies of interest. A recent Best Practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic. 
The uncertainty of a free energy estimate is limited however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation. As a consequence, the variance in free energy afforded by any estimator will always be an underestimate of the true uncertainty.
In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above. The user may decide to choose best practices based on their need for the calculations carried out. If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed. As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling scheme.

@@

>>>>

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Karwounopoulos2024}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Schoeller2023}
\end{itemize}

@@

\paragraph{Have you considered stereoisomers and enantiomers?}
Congeneric series can contain stereoisomers or enantiomers, which can bind very differently, resulting in large errors if treated incorrectly. For racemates, the relative abundance of each stereoisomer is normally not known, however when there is only one stereocenter in the molecule it can often be fair to assume that the substance is a 50\%:50\% mixture of the two. If one enantiomer is active (eutomer) and the other is inactive (distomer), this corresponds to the effective concentration of your bioactive molecule being half of what is expected in the assay, which corresponds to a ~0.2 pIC50 loss in potency relative to the pure eutomer being tested, which could be captured within the associated error calculations.\cite{h2011significance} If possible, both enantiomers could be considered computationally, however if a series contains multiple stereocenters, or multiple compounds within the series have stereocenters, this may be computationally restrictive. If compounds contain multiple undefined stereoecenters, the error and complication of understanding predictions may become prohibitive. Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain. If enumerating unknown stereocenters is not tractable, modeling using just the conformation that best fits the active site is reasonable, however this clearly introduces potential for larger errors compared to experiment. Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer. Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.

\paragraph{Are there cofactors expected to be present?}
Crystallographic structures can contain cofactors, which may be close to the ligand series of interest, and understanding if they are required or not for simulation can be an interdisciplinary effort between the modeller, crystallographer and bioligists to establish if they are a crystallographic artifact, or if they are likely to be present in the assay. In addition to both the crystallographic and assay conditions, what is most relevant \textit{in vivo} might be a third, equally difficult consideration. Following this, if a cofactor is considered to be included in the calculations, it is worth considering the appropriate forcefield for the cofactors. Cofactors, in particular metal ions can be challenging\cite{mobley2017predicting, guven2024protocols}. If the congeneric series of ligands is expected to interact with the cofactors differently, such as metal ions in different charge states, or to displace the cofactor for some of the molecules then this is likely beyond current modelling techniques. If there is still uncertainty around cofactor binding, running free energy calculations for a set of molecules both with and without the cofactor and comparing the relative rank-ordering of molecules could provide a sanity check.

@@

\subsection{Free energies of hydration or partition coefficients}
\tocsubsectioncomment{Steering Meeting: @Volunteer to expand coverage of free energies of hydration/ partition coefficients to make coverage of applications more equal.}
\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. However, in these cases, a careful choice of forcefield for the organic solvent model, as well as water model is essential. See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices. And, while sampling problems might seem to be a non-issue for small molecules, this is not always the case; e.g. even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{klimovich2010predicting, lim2019assessing}.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms and a hydroxyl group bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final states; see below for further details. In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other). %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond. Other force field parameters are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.

The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$. Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems. Giese et al. \cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}).

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors. \cite{ries2024Kartograf}

@@

In general, there are many ways on how an MCSS can be calculated. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

SIS based methods were shown to struggle with 3D properties of the molecules that need to be mapped, like for example stereochemistry. Therefore, some packages developed a post-mapping step, where they filtered for atom distances, and, therefore, added 3D information into the mapping.
Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}. The SIS MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.

Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules. These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g. stereochemistry or binding mode differing).
The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned. Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discussed below)|.
Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.

In any case, visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode. If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.

@@

While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}. Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule. However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom. Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.\ referred to this as “flapping”. Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed. In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small. For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017, bieniek2021}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task. In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}. Measurement of the sensitivity of individual edges in the thermodynamic graph to the enforcement of cycle closure constraints can help to identify problematic alchemical transformations~\cite{giese2025transferability}. In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them. Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses. However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues. Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear. Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.

@@

\subsection{Uncertainty estimation}
\tocsubsectioncomment{To be split between single/ multiple papers. @Agastya P Bhati/ Peter Coveney to expand. Cycle closure etc. to be moved to second paper.}
\tocsubsectioncomment{@Volunteer: Add discussion of bootstrapping vs t-based confidence intervals for repeats of single calculations (both bad options)?}
\label{subsec:uncertainty}
It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to obtain an estimate of uncertainty of the obtained value for the free energies of interest. A recent Best Practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic. 
The uncertainty of a free energy estimate is limited however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation. As a consequence, the variance in free energy afforded by any estimator will always be an underestimate of the true uncertainty.
In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above. The user may choose best practices based on their need for the calculations carried out. If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed. As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling and analysis scheme.

@@

More broadly, although several enhanced sampling strategies have been proposed to address configurational sampling bottlenecks in alchemical simulations, widely accepted best practices for detecting and resolving such challenges in a systematic and automated fashion remain elusive.
